Temporary: An Analysis of Programming Students’ Interaction Prompts with Generative AI: A Metacognitive Perspective

### 2024-09-30 CSEDU2025/Paper.tex
update

The study guided by \cite{chan23} revealed that both undergraduate and
postgraduate students exhibit positive attitudes toward the use of GenAI in
teaching and learning, noting that it enhances the depth of their thinking and
understanding. Additionally, a systematic review conducted by \cite{Lo24}
demonstrated that students could effectively learn from ChatGPT, leading to
improved understanding and academic achievement. Besides that, it was also
observed that ChatGPT enables students to regulate their learning pace
\citep{Baha24} and can potentially support students’ self-regulated learning,
especially for those with previous technical and disciplinary knowledge
\citep{Xia23}.

On the other hand, researchers have concerns about the impact of these tools on
students. The systematic review conducted by \cite{Murillo23} indicated that
ChatGPT use might lead to an overreliance on the tool. \cite{chan23} also noted
that this reliance could result in a decrease in critical thinking, as students
might make decisions based solely on the information provided by ChatGPT.

The student’s confidence is justified, as demonstrated by \cite{Puryear22}, who
found that GitHub Copilot can generate solutions for student assignments with
accuracy rates ranging from 68\% to 95\%. However, this also raises concerns
that students may become overly reliant on GenAI tools, potentially bypassing
the opportunity to develop a deeper understanding of underlying concepts.
\cite{cai23} further highlighted overdependence and diminished intellectual
engagement as significant drawbacks associated with the use of ChatGPT in
learning.

Regardless of teachers' preferences or beliefs, a preliminary surveys performed
by \cite{Dickey24} revealed that at least 54.5\% of students are already using
GenAI for homework. This highlights the need to increase the understanding of
how students are interacting with these tools and how they can be used to
enhance learning. Furthermore, a review organized by \cite{Lo24} emphasized the
necessity for extended studies and objective measures to gather more robust
evidence on the use of GenAI tools in education.

The primary objective of this study is to analyze the interactions between
students in a Java programming course and an educational bot powered by
ChatGPT 3.5, enhanced with Retrieval-Augmented Generation (RAG) capabilities,
within the context of two Metacognitive strategies: Interleaving and Spacing.
To achieve this objective, we formulated three research questions:

@@

However, it is crucial to consider how GenAI tools impact students across
various demographic groups, academic disciplines, cultural backgrounds, and
levels of prior experience \citep{catalan21} \citep{neo22}. Consequently, this
study is limited to a specific group of students and focuses on the use of a
single GenAI tool. As a result, the findings may not be generalizable to other
populations or tools.

@@

TODO: Work in progress

@@

About the first research question: \textit{What was the distribution pattern of
the classified messages?.}

According to Figure \ref{fig:graph1} messages classified as
Conceptual Question account for 32.1\% of the total. Often, the sequence of
Conceptual Question begins with a request for a Complete Solution or Code
Snippet, which then evolves into a series of Conceptual Question, as seen in
conversation 2 of Figure \ref{fig:graph2}. In other cases, the conversation
consists entirely of messages classified as Conceptual Question, such as in
conversation 10 of Figure \ref{fig:graph2}. This type of message indicates that
students recognize they have not yet fully mastered a concept and are seeking to
better understand it. In this context, the student realizes that solving the
problem requires understanding the concepts, thereby applying an active learning
strategy.

As shown in Figure \ref{fig:graph1}, approximately 30.9\% of the messages were
classified as Debugging Help. These prompts often begin with a request for a
Complete Solution or Code Snippet. Following this, students typically shift
from a passive to a more active learning strategy, as illustrated by the start
of conversation 12 in Figure \ref{fig:graph2}. Debugging Help messages represent
a form of active learning, where students likely focus on gaining a practical
understanding of the code, identifying errors, and improving their debugging
skills for future tasks.

Since most of the messages are located in categories where students are
notoriously engaged in more active in their learning (Conceptual Question and
Debugging Help) corroborating with the findings of \cite{Ghimire24} which showed
that the prompts are also localized in the same categories

Requests for the bot to generate a Code Snippet or Complete Solution accounted
for 10.5\% of student messages. These prompts reflected a desire for more
direct answers; however, it was observed that, after receiving these solutions,
many students transitioned to a more active approach in seeking to understand
the code or the underlying concepts. Thus, it can be inferred that prompts
within the Code Snippet or Complete Solution category were often used as a
starting point for more in-depth study.

Multiple-choice exercise resolutions accounted for 16.9\% of the responses.
These prompts typically involved students submitting one or more exercises to
CharlieBot and requesting solutions. We identified sequences of multiple-choice
questions, as illustrated by conversations 8, 9, and 15 in \ref{fig:graph2}.
This less active learning approach can occur for various reasons, ranging from
the belief that reviewing solutions to similar questions enhances learning, to
issues such as negligence, poor planning, or other possible factors.

Student corrections to the bot’s responses accounted for 2.6\% of the
interactions. In most cases, these corrections occurred when the student already
had the correct answer to an exercise and identified an error in the bot’s
response, as illustrated in conversations 4 and 9 in Figure \ref{fig:graph2}.
These corrections highlight that, although the bot provides quick feedback, it
is not always accurate, and its errors can be difficult for students to
identify.

>>>>

The study guided by \cite{chan23} revealed that both undergraduate and
postgraduate students exhibit positive attitudes towards the use of GenAI in
teaching and learning, noting that it improves the depth of their thinking and
understanding. Furthermore, a systematic review conducted by \cite{Lo24}
demonstrated that students could effectively learn from ChatGPT, leading to
improved understanding and academic achievement. Besides that, it was also
observed that ChatGPT enables students to regulate their learning pace
\citep{Baha24} and can potentially support students’ self-regulated learning,
especially for those with previous technical and disciplinary knowledge
\citep{Xia23}.

However, researchers are concerned about the impact of these tools on
students. The systematic review conducted by \cite{Murillo23} indicated that
the use of ChatGPT could lead to an overreliance on the tool. \cite{chan23} also noted
that this dependency could result in a decrease in critical thinking, as students
could make decisions based solely on the information provided by ChatGPT.

The student’s confidence is justified, as demonstrated by \cite{Puryear22}, who
found that GitHub Copilot can generate solutions for student assignments with
accuracy rates ranging from 68\% to 95\%. However, this also raises concerns
that students may become overly dependent on GenAI tools, potentially bypassing
the opportunity to develop a deeper understanding of the underlying concepts.
\cite{cai23} further highlighted the overdependence and diminished intellectual
engagement as significant drawbacks associated with the use of ChatGPT in
learning.

Regardless of teachers' preferences or beliefs, a preliminary surveys performed
by \cite{Dickey24} revealed that at least 54.5\% of students are already using
GenAI for homework. This highlights the need to increase the understanding of
how students interact with these tools and how they can be used to
enhance learning. Furthermore, a review organized by \cite{Lo24} emphasized the
need for extended studies and objective measures to gather more robust
evidence on the use of GenAI tools in education.

The primary objective of this study is to analyze the interactions between
students in a Java programming course and an educational bot powered by
ChatGPT 3.5, enhanced with Retrieval-Augmented Generation (RAG) capabilities,
within the context of two Metacognitive strategies: Interleaving and Spacing.
To achieve this objective, we formulate three research questions:

@@

However, it is crucial to consider how GenAI tools impact students in
various demographic groups, academic disciplines, cultural backgrounds, and
levels of previous experience \citep{catalan21} \citep{neo22}. Consequently, this
study is limited to a specific group of students and focuses on the use of a
single GenAI tool. As a result, the findings may not be generalizable to other
populations or tools.

@@

TODO: Work in progress.

@@

About the first research question: \textit{What was the distribution pattern of
the classified messages?}

According to Figure \ref{fig:graph1} messages classified as
Conceptual Question account for 32.1\% of the total. Often, the sequence of
Conceptual Question begins with a request for a Complete Solution or Code
Snippet, which then evolves into a series of Conceptual Question, as seen in
conversation 2 of Figure \ref{fig:graph2}. In other cases, the conversation
consists entirely of messages classified as Conceptual Question, such as in
conversation 10 of Figure \ref{fig:graph2}. This type of message indicates that
students recognize they have not yet fully mastered a concept and are seeking to
better understand it. In this context, the student realizes that solving the
problem requires understanding of concepts, thus applying an active learning
strategy.

As shown in Figure \ref{fig:graph1}, approximately 30.9\% of the messages were
classified as Debugging Help. These prompts often begin with a request for a
Complete Solution or Code Snippet. Following this, students typically switch from
passive to more active learning strategy, as illustrated by the start
of conversation 12 in Figure \ref{fig:graph2}. Debugging Help messages represent
a form of active learning, where students likely focus on gaining a practical
understanding of the code, identifying errors, and improving their debugging
skills for future tasks.

Since most of the messages are located in categories where students are
notoriously engaged in being more active in their learning (conceptual question 
and problem solving help), corroborating the findings of \cite{Ghimire24} which 
showed that the questions are also localized in the same categories.

Requests for the bot to generate a Code Snippet or Complete Solution accounted
for 10.5\% of student messages. These prompts reflected a desire for more
direct answers; however, it was observed that, after receiving these solutions,
many students transitioned to a more active approach in seeking to understand
the code or the underlying concepts. Thus, it can be inferred that the prompts
within the Code Snippet or Complete Solution category were often used as a
starting point for a more in-depth study.

Multiple Choice exercise resolutions represented 16.9\% of the responses.
These prompts typically involved students submitting one or more exercises to
CharlieBot and requesting solutions. We identified sequences of multiple choice
questions, as illustrated by conversations 8, 9, and 15 in \ref{fig:graph2}.
This less active learning approach can occur for various reasons, ranging from
the belief that reviewing solutions to similar questions enhances learning, to
issues such as negligence, poor planning, or other possible factors.

Student corrections to bot responses accounted for 2.6\% of the
interactions. In most cases, these corrections occurred when the student already
had the correct answer to an exercise and identified an error in the bot’s
response, as illustrated in conversations 4 and 9 in Figure \ref{fig:graph2}.
These corrections highlight that, although the bot provides quick feedback, it
is not always accurate, and its errors can be difficult for students to
identify.
